\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-06-14 canonicitymodel.tex
Save.

We follow the reducibility argument from \citep{coquand2018canonicity,sterling2019algebraic} to construct the canonicity model. We will first specify the formulation of the most basic canonicity theorem.

@@

Canonicity is one of the basic criteria to consider a dependent type theory to be ``a programming language or as a computational foundation for mathematics''\footnote{See nlab explanation: \href{https://ncatlab.org/nlab/show/canonical+form}{Canonical Form}}.
% cite https://ncatlab.org/nlab/show/canonical+form here
We can even argue that, if this theorem is proven in a constructive meta-logic\footnote{QIIT does have computational content}, 
% citation needed
then by Curry-Howard Correspondence, this theorem provides a big-step interpreter for closed term of boolean type.

@@

Another helpful analogy of the canonicity model is that $"Tm"_p~\Gamma^C~"T"^C~"t"$ can be understood as $"t" \in \{$ Reducible Terms of type $T$ (and $T_p$) in context $\Gamma$ (and $\Gamma_p$) $\}$. The latter is more conventional and the former can be understood as type theoretic encoding of the latter\footnote{We usually use predicate to encode "belong to" relation in type theory}. From this perspective, this part of the canonicity model is quite similar to how conventional (proof-irrelevant) logical relation is carried out --
% cite https://arxiv.org/abs/1907.11133
using the terminology of the conventional formulation, $\Gamma_p$ will be the set of substitution into closed reducible terms, and we will prove $t[\gamma] \in T_p$, a set of closed reducible term of type $T$ when $\gamma \in \Gamma_p$.

@@

      $\quad$ for some $\goodTerm{\cdot}{o}{\cL(p_1\nu~\sigma)}$ and $\goodTerm{\cdot, \cC(p_1\nu'~\sigma)}{t}{p_2\nu~\sigma}$
    \item if $\goodTerm{\cdot}{t}{\Sigma A B}$ then $\goodTerm{\cdot}{t = (a, b)}{\Sigma A B}$ with $\goodTerm{\cdot}{a}{A}$ and $\goodTerm{\cdot}{b}{B[(id, a)]}$
    \footnote{We emphasize the last one because $\cCt t$ is a dependent pair}
  \end{itemize}
\end{theorem}

>>>>

We follow the reducibility argument from \citep{coquand2018canonicity,sterling2019algebraic, kaposi2019gluing} to construct the canonicity model. We will first specify the formulation of the most basic canonicity theorem.

@@

Canonicity is one of the basic criteria to consider a dependent type theory to be ``a programming language or as a computational foundation for mathematics''\footnote{See nlab explanation: \href{https://ncatlab.org/nlab/show/canonical+form}{Canonical Form}}.
We can even argue that, if this theorem is proven in a constructive meta-logic\footnote{QIIT does have computational content\citep{altenkirch2016type}}, 
then by Curry-Howard Correspondence, this theorem provides a big-step interpreter for closed term of boolean type.

@@

Another helpful analogy of the canonicity model is that $"Tm"_p~\Gamma^C~"T"^C~"t"$ can be understood as $"t" \in \{$ Reducible Terms of type $T$ (and $T_p$) in context $\Gamma$ (and $\Gamma_p$) $\}$. The latter is more conventional and the former can be understood as type theoretic encoding of the latter\footnote{We usually use predicate to encode "belong to" relation in type theory}. From this perspective, this part of the canonicity model is quite similar to how conventional (proof-irrelevant) logical relation is carried out\citep{skorstengaard2019introduction} --
% cite https://arxiv.org/abs/1907.11133
using the terminology of the conventional formulation, $\Gamma_p$ will be the set of substitution into closed reducible terms, and we will prove $t[\gamma] \in T_p$, a set of closed reducible term of type $T$ when $\gamma \in \Gamma_p$.

@@

      $\quad$ for some $\goodTerm{\cdot}{o}{\cL(p_1\nu~\sigma)}$ and $\goodTerm{\cdot, p_1\nu'~\sigma}{t}{p_2\nu~\sigma}$
    \item if $\goodTerm{\cdot}{t}{\Sigma A B}$ then $\goodTerm{\cdot}{t = (a, b)}{\Sigma A B}$ with $\goodTerm{\cdot}{a}{A}$ and $\goodTerm{\cdot}{b}{B[(id, a)]}$
    \footnote{We emphasize the last one because $\cCt t$ is a dependent pair}
  \end{itemize}
\end{theorem}

### 2022-06-14 standardmodel.tex
Save.

Once we have the syntax of our theory, the first question is if our theory is consistent, i.e., if we can syntactically derive bottom in our theory. We prove the consistency by extending the standard model from \citep{kaposi2017type, kaposi2019gluing}, where they formally resort to the concept of \textit{algebra} of QIIT, here we only consider inductively interpreting each syntax piece into naive set theory that is also respecting judgemental equalities\footnote{In our appendix, we still interpret into Agda component to be more precise and less error-prone, also we need the inductive facility from Agda.} The main idea of the proof is that: 1. we use $\denotesS{}$ to map each intrinsic well-typed syntax piece into a mathematical object; 2. we define proof-irrelevant proposition $\models_S$ for each syntactical $\vdash$, and derive compatibility lemma. We will mainly focus on the definition of $\denotesS{}$ and omit the verification of $\models_S$\footnote{Actually our 1 is acting exactly as an interpretation function, and in our agda formulation we only do 1 because 2 is automatic; here we make 2 explicit to make our formulation more accessible and uniform}, and also we omit the verification of equality (i.e. the judgementally-equal syntax needs to be mapped to the same data for well-defined-ness).

@@

Notice that $\denotesS{\goodSig{\Gamma}{s}{n}}$ and $\denotesS{\goodType{\Gamma}{\cC\sigma}{i}}$ is mutual recursively defined together -- because $\goodSealS{\Gamma}{f}{\sigma}{A}$ relies on $\cC$. What's more, the substitution law around these two are also proved together by induction.

@@

The verification of validity of this model is of course required but largely omitted and we refer the reader to the appendix. For example, to verify the well-defined-ness of this model (i.e. judgemental equality is modeled by equality inside the model). If we want to verify compatibility lemma.

Finally, notice that we interpret the bottom type using empty set, and thus we know it is not possible to derive $\cdot \vdash t : \bot$: otherwise we have $\cdot \models_S t : \bot$ which exactly means $\denotesS{t}(\star) \in \denotesS{\bot}(\star) = \emptyset$ 
and that is a contradiction.

>>>>

Once we have the syntax of our theory, the first question is if our theory is consistent, i.e., if we can syntactically derive bottom in our theory. We prove the consistency by extending the standard model from \citep{altenkirch2016type,kaposi2017type, kaposi2019gluing} and their formalization, where they formally resort to the concept of \textit{algebra} of QIIT, here we only consider inductively interpreting each syntax piece into naive set theory that is also respecting judgemental equalities\footnote{In our appendix, we still interpret into Agda component to be more precise and less error-prone, also we need the inductive facility from Agda.} The main idea of the proof is that: 1. we use $\denotesS{}$ to map each intrinsic well-typed syntax piece into a mathematical object; 2. we define proof-irrelevant proposition $\models_S$ for each syntactical $\vdash$, and derive compatibility lemma. We will mainly focus on the definition of $\denotesS{}$ and omit the verification of $\models_S$\footnote{Actually our 1 is acting exactly as an interpretation function, and in our agda formulation we only do 1 because 2 is automatic; here we make 2 explicit to make our formulation more accessible and uniform}, and also we omit the verification of equality (i.e. the judgementally-equal syntax needs to be mapped to the same data for well-defined-ness).

@@

Notice that $\denotesS{\goodSig{\Gamma}{s}{n}}$ and $\denotesS{\goodType{\Gamma}{\cC\sigma}{i}}$ is mutual recursively defined together -- because $\goodSealS{\Gamma}{f}{\sigma}{A}$ relies on $\cC$.

@@

The verification of validity of this model is largely omitted here and we refer the reader to the appendix. To verify our model preserve the judgemental equality, usually unpacking the definition is enough -- for example, to verify $\beta$-rule for linkage projection, we only need to see that $\denotesS{\goodTerm{\Gamma}{p_1\mu~(\mu^+~o~t)}{\cL\sigma}} = \denotesS{\goodTerm{\Gamma}{o}{\cL\sigma}}$. One exception is for $\cC$ and $\goodSig{\_}{\_}{n}$, the substitution law around these two are also proved together by induction because they were mutual-recursively defined together. The compatibility lemmas can be verified directly -- for example $\frac{\goodSigS{\Gamma}{\sigma}{n+1}\quad\goodTermS{\Gamma}{o}{\cL\sigma}}{\goodTermS{\Gamma}{p_1\mu~o}{\cL(p_1\nu~\sigma)}}$ is proved by simply unfolding the definition.

Since all the compatibility lemmas hold (by simply changing $\vdash$ into $\models$ in the syntactic rules then we would have all the compatibility lemmas\footnote{Apparently, in our (fake) Agda formulation, we don't need this extra step.}), we have constructed a model.
Finally, notice that we interpret the bottom type using empty set, and thus we know it is not possible to derive $\cdot \vdash t : \bot$: otherwise we have $\cdot \models_S t : \bot$ which exactly means $\denotesS{t}(\star) \in \denotesS{\bot}(\star) = \emptyset$ 
and that is a contradiction.
